Acceleration of Global Optimization Algorithm by Detecting Local Extrema Based on Machine Learning

### 2021-08-27 2021/Entropy/entropy.tex
some fix
fix accuracy

% Affiliations / Addresses (Add [1] after \address if there is only one affiliation.)
\address[1]{%
$^{1}$ \quad Department of Mathematical Software and Supercomputing Technologies, Lobachevsky University, 603950 Nizhny Novgorod, Russia; ilya.lebedev@itmm.unn.ru (I.L.), evgeny.kozinov@itmm.unn.ru (E.A.)}
%$^{2}$ \quad Affiliation 2; e-mail@e-mail.com}

@@

As a standard, let us consider a one-dimensional multiextremal optimization problem
\begin{equation}\label{uni_problem}
\varphi^\ast = \varphi(x^\ast)=\min{\left\{\varphi(x):x\in \left[0,1\right] 
\right\}}
\end{equation}
with an objective function satisfying the Lipschitz condition.

@@

In accordance with the algorithm, the first two trials are carried out at the boundary points of the segment $[0,1]$, т.е. $x^0=0,\;x^1=1$. 
At these points, the values of the objective function  $z^0=\varphi(x^0),\;z^1=\varphi(x^1)$ are calculated and the counter value is set to $k=1$. 
The point of the next trial $x^{k+1}, k\geq 1,$ is selected in accordance with the following procedure.

 Step 1. Renumber  (starting at 0) the points $x^i,\:0\leq i\leq k$, of the trials conducted  in ascending order of the coordinate, i.e.
\begin{equation}\label{xt}
0=x_0<x_1<\ldots <x_{k}=1.
\end{equation} 
Associate the values of the objective function $z_i=\varphi(x_i), \; 0\leq i\leq k,$ to the points $x_i, \; 0\leq i\leq k$, at which these values were calculated.

@@

The global optimization methods discussed above were compared when solving 100 problems from the  $F_{SH}$ and $F_{HL}$ classes. The problem will be considered correctly solved if after stopping the method by accuracy (i.e. when the length of the current search interval becomes less than $\epsilon$) the current estimate of the optimum  $x_k^*$ lies in the $\epsilon$-neighborhood of the known solution of the problem  $x^*$, i.e. if the condition  $|x^*-x_k^*| \leq \epsilon$ is satisfied.

Tables \ref{table:average_Shekel} and \ref{table:average_Hill} show the number of search trials that, on average, were required to minimize the Shekel and Hill functions with different search accuracy  $\epsilon$. The number of unsolved problems is indicated in parentheses. The experimental results show that with a rough solution to the problem, all methods show similar results in terms of the number of trials, while with a high solution accuracy, the GSA-DT algorithm requires 2 times fewer trials than its prototype. At the same time, GSA-DT outperforms the DIRECT method both in the average number of search trials and in the number of correctly solved problems. In particular, if we use the accuracy $\epsilon = 10^{-2}$, then the DIRECT method stops too early and does not find a global solution to many problems. 
Therefore, in further experiments in which multidimensional problems are solved, we will not use DIRECT for comparison, since when solving multidimensional problems with stopping by accuracy, this method will provide a correct solution to no more than  $50\%$ of problems.

@@

As an example, consider the operation of the GSA and GSA-DT algorithms when solving one of the two-dimensional problems generated by the GKLS generator. 
The level lines of the objective function shown in Figures \ref{fig3} and \ref{fig4}, indicate the presence of ten local extrema. 
When starting the algorithms, the same parameters were used:  $r=3.0$ from (\ref{R}) and $\epsilon = 10^{-2}$ in the stopping criterion of the method. 
Black dots in Figures \ref{fig3} and \ref{fig4} show the points of the search trials performed by the methods in the process of solving the problem. In this case, the GSA algorithm required 247 trials, while the GSA-DT algorithm took 138 trials.
The red dot in the figures marks the exact solution of the problem, and the yellow dot indicates the best approximation found by the algorithm. 
Green dots in Figure \ref{fig4} indicate trials performed as part of a local search. 
This graph demonstrates that using decision trees to identify areas of attraction of local minima removes the problem of accumulation of test points in the region of local extrema inherent in the original global search algorithm.

@@

Tables \ref{table:average_GKLS01} and \ref{table:average_GKLS002} show the average number of trials required by the GSA and GSA-DT methods to correctly solve all problems with an accuracy of $\epsilon = 10^{-2}$ и $\epsilon = 2 \cdot 10^{-3}$ respectively.

@@

\begin{specialtable}[H] 
	\caption{Solving GKLS problems with an accuracy  $\epsilon = 10^{-2}$}\label{table:average_GKLS01}
	\center
\begin{tabular}{cccc}
\toprule
        & $N=2$ & $N=3$  & $N=4$    \\
\midrule
GSA     &  937  &  12716 &  206869  \\
GSA-DT  &  653  &  9204  &  156190  \\
\bottomrule
\end{tabular}
\end{specialtable}

\begin{specialtable}[H] 
	\caption{Solving GKLS problems with an accuracy $\epsilon = 2 \cdot 10^{-3}$}\label{table:average_GKLS002}
	\center
\begin{tabular}{cccc}
\toprule

@@

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\authorcontributions{Conceptualization and methodology, K.B.; software and validation, I.L., E.A.; formal analysis, K.B.; investigation, I.L.; data curation, I.L.; writing --- original draft preparation, K.B.; writing- -- review and editing, K.B.; visualization, I.L.; funding acquisition, K.B. All authors have read and agreed to the published version of the manuscript.}

>>>>

% Affiliations / Addresses (Add [1] after \address if there is only one affiliation.)
\address[1]{%
$^{1}$ \quad Department of Mathematical Software and Supercomputing Technologies, Lobachevsky University, 603950 Nizhny Novgorod, Russia; ilya.lebedev@itmm.unn.ru (I.L.), evgeny.kozinov@itmm.unn.ru (E.K.)}
%$^{2}$ \quad Affiliation 2; e-mail@e-mail.com}

@@

As a standard, let us consider a one-dimensional multiextremal optimization problem
\begin{equation}\label{uni_problem}
\varphi^\ast = \varphi(x^\ast)=\min{\left\{\varphi(x):x\in \left[a,b\right], a<b, a,b \in R
\right\}}
\end{equation}
with an objective function satisfying the Lipschitz condition.

@@

In accordance with the algorithm, the first two trials are carried out at the boundary points of the segment $[a,b]$, i.e. $x^0=a,\;x^1=b$. 
At these points, the values of the objective function  $z^0=\varphi(x^0),\;z^1=\varphi(x^1)$ are calculated and the counter value is set to $k=1$. 
The point of the next trial $x^{k+1}, k\geq 1,$ is selected in accordance with the following procedure.

 Step 1. Renumber  (starting at 0) the points $x^i,\:0\leq i\leq k$, of the trials conducted  in ascending order of the coordinate, i.e.
\begin{equation}\label{xt}
a=x_0<x_1<\ldots <x_{k}=b.
\end{equation} 
Associate the values of the objective function $z_i=\varphi(x_i), \; 0\leq i\leq k,$ to the points $x_i, \; 0\leq i\leq k$, at which these values were calculated.

@@

The global optimization methods discussed above were compared when solving 100 problems from the  $F_{SH}$ and $F_{HL}$ classes. The problem will be considered correctly solved if after stopping the method by accuracy (i.e. when the length of the current search interval becomes less than $\epsilon \cdot \left|b-a\right| $) the current estimate of the optimum  $x_k^*$ lies in the $\epsilon$-neighborhood of the known solution of the problem  $x^*$, i.e. if the condition  $|x^*-x_k^*| \leq \epsilon \cdot \left|b-a\right|$ is satisfied.

Tables \ref{table:average_Shekel} and \ref{table:average_Hill} show the number of search trials that, on average, were required to minimize the Shekel and Hill functions with different search accuracy  $\epsilon$. The number of unsolved problems is indicated in parentheses. The experimental results show that with a rough solution to the problem, all methods show similar results in terms of the number of trials, while with a high solution accuracy, the GSA-DT algorithm requires 2 times fewer trials than its prototype. At the same time, GSA-DT outperforms the DIRECT method both in the average number of search trials and in the number of correctly solved problems. In particular, if we use the accuracy $\epsilon = 10^{-2}\left|b-a\right|$, then the DIRECT method stops too early and does not find a global solution to many problems. 
Therefore, in further experiments in which multidimensional problems are solved, we will not use DIRECT for comparison, since when solving multidimensional problems with stopping by accuracy, this method will provide a correct solution to no more than  $50\%$ of problems.

@@

As an example, consider the operation of the GSA and GSA-DT algorithms when solving one of the two-dimensional problems generated by the GKLS generator. 
The level lines of the objective function shown in Figures \ref{fig3} and \ref{fig4}, indicate the presence of ten local extrema. 
When starting the algorithms, the same parameters were used:  $r=3.0$ from (\ref{R}) and $\epsilon = 10^{-2}\left|b-a\right|$ in the stopping criterion of the method. 
Black dots in Figures \ref{fig3} and \ref{fig4} show the points of the search trials performed by the methods in the process of solving the problem. In this case, the GSA algorithm required 247 trials, while the GSA-DT algorithm took 138 trials.
The red dot in the figures marks the exact solution of the problem, and the yellow dot indicates the best approximation found by the algorithm. 
Green dots in Figure \ref{fig4} indicate trials performed as part of a local search. 
This graph demonstrates that using decision trees to identify areas of attraction of local minima removes the problem of accumulation of test points in the region of local extrema inherent in the original global search algorithm.

@@

Tables \ref{table:average_GKLS01} and \ref{table:average_GKLS002} show the average number of trials required by the GSA and GSA-DT methods to correctly solve all problems with an accuracy of $\epsilon = 10^{-2}\left|b-a\right|$ и $\epsilon = 2 \cdot 10^{-3}\left|b-a\right|$ respectively.

@@

\begin{specialtable}[H] 
	\caption{Solving GKLS problems with an accuracy  $\epsilon = 10^{-2}\left|b-a\right|$}\label{table:average_GKLS01}
	\center
\begin{tabular}{cccc}
\toprule
        & $N=2$ & $N=3$  & $N=4$    \\
\midrule
GSA     &  937  &  12716 &  206869  \\
GSA-DT  &  653  &  9204  &  156190  \\
\bottomrule
\end{tabular}
\end{specialtable}

\begin{specialtable}[H] 
	\caption{Solving GKLS problems with an accuracy $\epsilon = 2 \cdot 10^{-3}\left|b-a\right|$}\label{table:average_GKLS002}
	\center
\begin{tabular}{cccc}
\toprule

@@

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\authorcontributions{Conceptualization and methodology, K.B.; software and validation, I.L., E.K.; formal analysis, K.B.; investigation, I.L.; data curation, I.L.; writing --- original draft preparation, K.B.; writing- -- review and editing, K.B.; visualization, I.L.; funding acquisition, K.B. All authors have read and agreed to the published version of the manuscript.}
